### 2023-02-25 calculus.tex
Add more figures and discussion of marking calculus rules

\section{Motivating Examples}
In this section, we motivate a formal calculus for bidirectional type error localization through a
number of examples.

% TODO Talk about idea that bidirectional typing makes error localization easy?

\subsection{Unbound Variables}
% A simple example of an error that may occur in programs is the use of an unbound variable.

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. For example, in an expression such as
$\EPlus{\EMV_1}{\EMV_2}$, both $\EMV_1$ and $\EMV_2$ should have the type $\TNum$. In a
bidirectional system, $\EMV_1$ and $\EMV_2$ would be analysed against $\TNum$.

@@

\section{Marking Calculus}
\label{sec:calculus}

The previous section introduced different error localization cases by example. Now, we formally
introduce a \emph{marking calculus} for bidirectional type error localization on a lambda calculus
extended with numbers, conditionals, and let bindings, given by a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} ($\EMName$), which are the surface-level expression syntax.
    Types classify unmarked expressions according to a standard bidirectional gradual typing
    semantics. Note that a non-gradual system may be recovered straightforwardly by removing the
    unknown type and replacing type consistency with equality (discussed in
    \cref{sec:calculus-non-gradual}).

@@

  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}

In the interest of space, this overview focuses on the ``interesting'' cases; the whole system,
alongside the metatheorems discussed have been mechanized in the Agda proof assistant (discussed in
\cref{sec:calculus-agda}).

@@

\Cref{fig:calculus-syntax-uexp} gives the syntax of types, $\TMV$, and unmarked expressions, $\EMV$,
which extend the simply typed lambda calculus with $\TNum$ and $\TBool$ base types of numbers and
booleans. The number literal corresponding to the mathematical number $n$ is given by $\ENumMV$, and
there is a single addition operation on numeric expressions. $\ETrue$ and $\EFalse$ correspond to
the boolean values $\textsf{true}$ and $\textsf{false}$, and the branching expression
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else form.

In addition, types are extended with an \emph{unknown type}, $\TUnknown$, which

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \TMName  & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \EMName  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \mid \ELet{x}{\EMV}{\EMV} \\
             &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
                       \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
  \end{array}\]
  \caption{Syntax of types and unmarked expressions.}
  \label{fig:calculus-syntax-uexp}
\end{figure}

Unmarked expressions are governed by a standard bidirectional gradual type system, as given in
\Cref{fig:calculus-typing-uexp}.

% TODO Not sure how explicitly we want to spell out the type system, in lieu of background
% section on bidirectional typing?

% \input{figs/calculus-type-judgments.tex}
% \input{figs/calculus-typing-uexp.tex}

@@

Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with explicit \emph{error
markings}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three
such forms, which mirror the examples above:
%
\begin{itemize}
  \item $\ECUnbound{x}$, denoting the usage of an unbound variable.

  \item $\ECInconType{\ECMV}$, which occurs when the synthesized type of $\ECMV$ is inconsistent
    with the required type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, which occurs when the types of the two branches
    $\ECMV_2$ and $\ECMV_3$ are inconsistent.
\end{itemize}
%
Additional forms may arise as required by the language.

@@

@@

\input{figs/calculus-typing-mexp.tex}

@@

\subsection{Well-Formedness}
\label{sec:calculus-wellformedness}

@@

>>>>

\section{Motivating Examples}
\emph{Bidirectional typing} combines type checking, which determines if a program satisfies a given
type, and type synthesis, which generates a type from the program. This gives a simple algorithmic
system in which type information is recursively propagated through terms and allows for local type
inference.

% TODO talk about idea that bidirectional typing makes error localization easy? i.e. why choose
% bidirectional?

Often, programmers are working with program states that are syntactically well-formed but not
well-typed. However, formal typing semantics usually only give meaning to well-typed programs,
leaving ill-typed ones meaningless. This is insufficient for editor services and language servers,
which give users feedback on where and why a program is ill-typed. Traditionally, this has been
performed in an \emph{ad hoc} manner [citations?].

In this section, we motivate a formal calculus for bidirectional type error localization by
exploring different kinds of type errors that may occur.

% TODO state goals: prefer fine-grained over coarser blaming/marking

\subsection{Unbound Variables}
A simple and common example of an error that may occur in programs is the use of an unbound
variable. In this case, editor services often indicate visually that such usages are invalid.

% TODO image of an editor for this example? not sure what else to say about this

\subsection{Inconsistent Types}
Often, the type an expression \emph{should} be is known. Both operands of an addition expression
$\EPlus{\ETrue}{\ENumMV}$, for example, should have the type $\TNum$. In a bidirectional system,
$\ETrue$ and $\ENumMV$ would be analysed against $\TNum$, in which the former check would fail:
$\ETrue$ synthesizes $\TBool$, which is inconsistent with $\TNum$.

% An editor may indicate such an error in different ways. Two obvious ones are considered below: (1)
% indicating an error on $\ETrue$, or (2) indicating an error on the entire expression
% $\EPlus{\ETrue}{\ENumMV}$.
% \[%
  % \EPlus{\ETrue}{\ENumMV} \qquad \EPlus{\ETrue}{\ENumMV}
% \]%
% The former ``blames'' the error on the operand,

% Since the ``fault'' of the error

% TODO cite real editor/language?

Since expressions typed in analytic mode are checked to satisfy a correct, known type, they are
given ``blame'' when ill-typed.

@@

\subsection{Correctness}
\begin{theorem}[name=Marking Totality]
  Any syntactically well-formed program may be marked for errors.
\end{theorem}

In the next section, we will formally develop

\section{Marked Lambda Calculus}
\label{sec:calculus}

The previous section surveyed different type error cases. Now, we formally introduce a \emph{marking
calculus} for bidirectional type error localization on the gradually typed lambda calculus above.
This comes as a two-layer system:
%
\begin{itemize}
  \item \textbf{Unmarked expressions} ($\EMName$), which compose the surface-level syntax. Types
    classify unmarked expressions according to a standard bidirectional gradual typing semantics.

@@

  \item \textbf{Mark insertion}, which \textbf{marks} any unmarked expression into a marked
    expression, inserting error markings where appropriate.
\end{itemize}
%
That is, we extend the ordinary typing semantics of a language with a secondary ``marked'' language
and define a mapping from ``unmarked'' programs to ``marked'' programs. While the semantics that
follow concern a gradually typed language, a non-gradual system may be recovered straightforwardly
by removing the unknown type and replacing type consistency with equality.

In the interest of space, this overview focuses on the ``interesting'' cases. All rules and
metatheorems may be found in the supplementary appendix, alongside a complete mechanization in the
Agda proof assistant (discussed in \cref{sec:calculus-agda}).

@@

We consider a gradually typed lambda calculus extended with numbers and conditionals, whose syntax
and bidirectional typing are given in \cref{fig:calculus-syntax-example} and
\cref{fig:calculus-typing-example}, respectively. The base types $\TNum$ and $\TBool$ classify
number and boolean expressions. The number literal corresponding to the mathematical number $n$ is
given by $\ENumMV$, and there is a single addition operation on numeric expressions. $\ETrue$ and
$\EFalse$ correspond to the boolean values $\textsf{true}$ and $\textsf{false}$, and
$\EIf{\EMV_1}{\EMV_2}{\EMV_3}$ gives an if-else structure. In addition to these forms, $\TUnknown$
indicates an unknown type.

\begin{figure}[htbp]
  \[\begin{array}{rrcl}
    \textsf{Type} & \TMV  & \Coloneqq & \TUnknown \mid \TNum \mid \TBool \mid \TArrow{\TMV}{\TMV} \\
    \textsf{Exp}  & \EMV  & \Coloneqq & x \mid \ELam{x}{\TMV}{\EMV} \mid \EAp{\EMV}{\EMV} \\
         &       & \mid         & \ENumMV \mid \EPlus{\EMV}{\EMV}
                   \mid           \ETrue \mid \EFalse \mid \EIf{\EMV}{\EMV}{\EMV}
  \end{array}\]
  \caption{Syntax of the extended gradually typed lambda calculus}
  \label{fig:calculus-syntax-uexp}
\end{figure}

Bidirectional typing employs two mutually defined judgments. \emph{Type synthesis}, written
$\ctxSynType{\ctx}{\EMV}{\TMV}$, establishes that, under the typing context $\ctx$, the expression
$\EMV$ synthesizes or infers the type $\TMV$. \emph{Type analysis}, written
$\ctxAnaType{\ctx}{\EMV}{\TMV}$, states that the expression $\EMV$ may appear where an expression of
type $\TMV$ is expected. The \emph{subsumption} rule (ASubsume) states that if an expression
synthesizes a type, it may be analysed against that type or any \emph{consistent} type.

Note that the subsumption rule requires that $\EMV$ be ``subsumable'', a judgment defined on all
forms except lambda abstractions and if-else expressions. This restricts the usage of subsumption to
a rule of ``last resort'', admitting the unicity of typing.

% \begin{theorem}[name=Unmarked Expression Typing Unicity] \
  % If $\ctxSynType{\ctx}{\EMV}{\TMV}$ and $\ctxSynType{\ctx}{\EMV}{\TMV'}$, then $\TMV = \TMV'$.
% \end{theorem}

\input{figs/calculus-typing-uexp}

The \emph{consistency} judgment $\consistent{\TMV}{\TMV'}$ that appears in the subsumption rule
defines a reflexive and symmetric (but not transitive) relation between types. The unknown type is
additionally consistent with any other type, and arrow types are consistent if their components are.

$\matchedArrow{\TMV}{\TMV_1}{\TMV_2}$ gives that $\TMV$ is a \emph{matched arrow type}, which means
that it may be considered an arrow type. It is purely a technical mechanism to avoid duplication of
rules related to arrow types.

\input{figs/calculus-type-judgments}

@@

Marked expressions, $\ECMV$, extend the syntax of unmarked expressions with \emph{explicit error
marks}. As given in the syntax definition of \cref{fig:calculus-syntax-mexp}, there are three such
forms, which mirror the examples discussed previously:
%
\begin{itemize} % TODO not the snappiest names
  \item $\ECUnbound{x}$, the \emph{unbound variable mark}, denotes the usage of an unbound variable.

  \item $\ECInconType{\ECMV}$, the \emph{inconsistent types mark}, occurs when the synthesized type
    of $\ECMV$ is inconsistent with the required type.

  \item $\ECInconBr{\ECMV_1}{\ECMV_2}{\ECMV_3}$, the \emph{inconsistent branches mark}, corresponds
    to an if-else expression where the types of the two branches $\ECMV_2$ and $\ECMV_3$ are
    inconsistent.
\end{itemize}
%
Additional forms may arise as required by the language.

% TODO examples of such?

@@

The unmarked version of a marked expression $\ECMV$ may be recovered through \emph{mark erasure},
given by $\erase{\ECMV}$. As in \Cref{fig:calculus-mark-erasure}, this total function may be defined
inductively in a straightforward manner. Observe that erasing the inconsistent branches mark gives
an if-else expression with sub-expression marks recursively erased.

\input{figs/calculus-mark-erasure}

@@

\input{figs/calculus-typing-mexp}

@@

\subsection{Correctness}
\label{sec:calculus-correctness}

% Totality

% Well-formedness conditions

@@

% TODO extension to non-typing semantics?
